---
title: "IMC data analysis workflow"
author: "Tess and Tural"
date: "University of Bern, DBMR, Visceral and Transplantation Surgery lab"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Dimensionality reduction 

```{r setup, include=FALSE}
#Setup knitr, incuding path
knitr::opts_chunk$set(echo = TRUE)

#Important - this path is relative to your dataset, set working directory
knitr::opts_knit$set(root.dir = "C:/Users/IMC Platform laptop/Users2/Tess/SteinbockCellSegmentation/221130_haibin/230515_workflowr_haibin")

#Set cache
knitr::opts_chunk$set(autodep = TRUE, cache = TRUE, cache.lazy = FALSE)

#To make figures
knitr::opts_chunk$set(dev = "png")
```

```{css, echo=FALSE}
.scroll-100 {
  max-height: 100px;
  overflow-y: auto;
  background-color: inherit;
}
```

### Load packages 

```{r load-libs, message=FALSE}
library(SpatialExperiment)
library(Rphenograph)
library(igraph)
library(dittoSeq)
library(viridis)
library(bluster)
library(BiocParallel)
library(ggplot2)
library(scran)
library(CATALYST)
library(kohonen)
library(ConsensusClusterPlus)
library(patchwork)
library(pheatmap)
library(gridExtra)
library(cytomapper)
library(Rphenoannoy)
library(SingleCellExperiment)
library(caret)
library(ggridges)
library(openxlsx)
library(stringr)
library(imcRtools)
library(cytomapper)
library(SpatialExperiment)
library(cytomapper)
library(scater)
library(patchwork)
library(cowplot)
library(viridis)
library(scuttle)
library(CATALYST)
library(ComplexHeatmap)
library(circlize)
library(tidyverse)
library(gtools)
library(RColorBrewer)
library(scuttle)
```

### Load data

```{r load data}
spe <- readRDS(file.path("output", "spe_ca.rds"))

# Sample cells
#cur_cells <- sample(seq_len(ncol(spe)), 2000)
```

### Add metadata 

```{r metadata}
#Metadata table loading
meta <- read.xlsx("./data/sample_metadata_shortened.xlsx")

# Set patient ID and ROI and condition names
spe$patient_id <- meta$patient_id[match(spe$sample_id, meta$sample_id)]
spe$patient_id <- factor(spe$patient_id)

#Factor ordering the ROI IDs for better plotting
spe$ROI <- meta$ROI[match(spe$sample_id, meta$sample_id)]
spe$ROI <- factor(spe$ROI, levels = mixedsort(unique(spe$ROI)))

spe$condition <- meta$condition[match(spe$sample_id, meta$sample_id)]
spe$annotation <- meta$annotation[match(spe$sample_id, meta$sample_id)]

spe@metadata$meta_final <- meta

```

### Check 2 - metadata

```{r checkmetadata, class.output="scroll-100"}
unique(spe$patient_id)
unique(spe$ROI)
unique(spe$condition)
unique(spe$annotation)
```


### Dimensionality reduction: annotated cells

```{r celltype overlay UMAP}
dittoDimPlot(spe, 
             var = "celltype",
             reduction.use = "UMAP", 
             size = 0.2,
             do.label = TRUE) +
    ggtitle("IMCA annotated cell types overlaid on UMAP")

```

#Celltype UMAP split by condition

```{r celltype by condition, fig.width= 10, fig.height= 12}
#Sample 10% of the dataset
spe_small <- spe[, sample(ncol(spe), round(ncol(spe)/10))]

p1 <- dittoDimPlot(spe_small, 
             var = "celltype",
             split.by = "condition",
             reduction.use = "UMAP", 
             size = 0.6,
             do.label = FALSE)

p2 <- dittoBarPlot(spe, var = "celltype", group.by = "condition")
p3 <- dittoBarPlot(spe, var = "condition", group.by = "celltype")

plot_grid(p1, p2, p3,
          ncol = 1)

```

#Double split barplots

```{r celltype by condition and annotation, fig.width= 10, fig.height= 12}
plotdata <- as.data.frame(colData(spe)[, c("celltype", "annotation", "condition")])
plotdata <- as.data.frame(table(plotdata)) 

ggplot(plotdata, aes(x = annotation, y = Freq, fill = condition)) + 
  geom_col() +
  facet_grid( ~ celltype)

ggplot(plotdata, aes(x = annotation, y = Freq, fill = condition)) + 
  geom_col() +
  facet_wrap( ~ celltype, scales = "free", ncol = 14)

```

#Celltype UMAP split by annotation

```{r celltype by annotation, fig.width= 10, fig.height= 12}

p1 <-dittoDimPlot(spe_small, 
             var = "celltype",
             split.by = "annotation",
             reduction.use = "UMAP", 
             size = 0.6,
             do.label = FALSE)

p2 <-dittoBarPlot(spe, var = "celltype", group.by = "annotation")
p3 <-dittoBarPlot(spe, var = "annotation", group.by = "celltype")

plot_grid(p1, p2, p3,
          ncol = 1)

```

#Celltype UMAP split by patient ID

```{r celltype by patientID, fig.width= 10, fig.height= 16}

p1 <-dittoDimPlot(spe_small, 
             var = "celltype",
             split.by = "donor_id",
             reduction.use = "UMAP", 
             size = 0.6,
             do.label = FALSE)

p2 <-dittoBarPlot(spe, var = "celltype", group.by = "donor_id")
p3 <-dittoBarPlot(spe, var = "donor_id", group.by = "celltype")

plot_grid(p1, p2, p3,
          ncol = 1)

```

#Celltype UMAP split by ROI

```{r celltype by ROI, fig.width= 10, fig.height= 16}

p1 <-dittoDimPlot(spe_small, 
             var = "celltype",
             split.by = "ROI",
             reduction.use = "UMAP", 
             size = 0.6,
             do.label = FALSE)

p2 <-dittoBarPlot(spe, var = "celltype", 
             group.by = "ROI", 
             x.reorder = mixedorder(metaLevels("ROI", spe)))
p3 <-dittoBarPlot(spe, var = "ROI", group.by = "celltype")

plot_grid(p1, p2, p3,
          ncol = 1)

```

# Target expression UMAPs

```{r UMAP target exprs, fig.width= 12, fig.height= 12}

# Define markers
markers <- c("E-cadherin", "Arginase-1", "CD19", "CD3e", "Pan-actin", "CD206", "HNFa", "a-SMA", "CD8a", "CD4", "CD11b", "Ly6-G", "CD31", "F4_80", "Ki-67", "CD44", "CD74")

#Multiple marker plots
plot_list <- multi_dittoDimPlot(spe_small, 
                                var = markers, 
                                assay = "exprs", 
                                reduction.use = "UMAP",
                                list.out = TRUE,
                                size = 0.2)

plot_list <- lapply(plot_list, function(x) x + scale_color_viridis())

plot_grid(plotlist = plot_list)
```

# Single target plots

```{r target overlay UMAP, eval=FALSE, include=FALSE}
#Single marker plots

#E-cadherin plot split by condition
p1 <- dittoDimPlot(spe_small, 
             var = "E-cadherin",
             split.by = "condition",
             assay = "exprs", 
             reduction.use = "UMAP", 
             size = 0.6,
             do.label = FALSE) + scale_color_viridis()

#E-cadherin plot split by annotation
p2 <- dittoDimPlot(spe_small, 
             var = "E-cadherin",
             split.by = "annotation",
             assay = "exprs",
             reduction.use = "UMAP", 
             size = 0.6,
             do.label = FALSE) + scale_color_viridis()

#E-cadherin plot split by patient ID
p3 <- dittoDimPlot(spe_small, 
             var = "E-cadherin",
             split.by = "donor_id",
             assay = "exprs",
             reduction.use = "UMAP", 
             size = 0.6,
             do.label = FALSE) + scale_color_viridis()

plot_grid(p1, p2, p3,
          ncol = 1)

```

# Get rid of useless channels

```{r filterdata}
#Exclude unwanted channels 

rowData(spe)$use_channel <- grepl("a-SMA|CD19|HNFa|CD11b|F4_80|CD44|CD3e|E-cadherin|Ly6-G|CD206|Arginase-1|CD4|CD74|CD8a|Ki-67|CD31|Pan-actin|HistoneH3|DNA1|DNA2", rownames(spe))
```

### Define colors

```{r set colors}

color_vectors <- list()

#For the ROI
ROI <- setNames(dittoColors(reps = 1)[seq_along(unique(spe$sample_id))], 
                unique(spe$ROI))
#For the patient ID
patient_id <- setNames(brewer.pal(length(unique(spe$patient_id)), name = "Set1"), 
                unique(spe$patient_id))
#For the sample ID
sample_id <- setNames(dittoColors(reps = 1)[seq_along(unique(spe$sample_id))], 
                unique(spe$sample_id))
#For the condition
condition <- setNames(brewer.pal(length(unique(spe$condition)), name = "Set2"), 
                unique(spe$condition))
#For the annotation
annotation <- setNames(brewer.pal(length(unique(spe$annotation)), name = "Set3"), 
                unique(spe$annotation))
#For cell type
annotation <- setNames(brewer.pal(length(unique(spe$annotation)), name = "Set3"), 
                unique(spe$annotation))

color_vectors$ROI <- ROI
color_vectors$patient_id <- patient_id
color_vectors$sample_id <- sample_id
color_vectors$condition <- condition
color_vectors$annotation <- annotation

metadata(spe)$color_vectors <- color_vectors
```

# Heatmap visualization

```{r IMCA cell types heatmap singlecell, fig.width= 10, fig.height= 10}

# Sample cells
cur_cells <- sample(seq_len(ncol(spe)), 40000)

dittoHeatmap(spe[, cur_cells], 
             genes = rowData(spe)$use_channel,
             assay = "exprs", scale = "none",
             heatmap.colors = viridis(100), 
             annot.by = c("celltype", "condition", "annotation", "patient_id"),
             annot.colors = c(dittoColors(1)[1:length(unique(spe$celltype))], metadata(spe)$color_vectors$condition, metadata(spe)$color_vectors$annotation, metadata(spe)$color_vectors$patient_id))
             
```


```{r heatmap_cellsaggregated, fig.width= 10, fig.height= 10}

# Define type markers
type_markers <- c("E-cadherin", "Arginase-1", "CD19", "CD3e", "Pan-actin", "CD206", "HNFa", "a-SMA", "CD8a", "CD4", "CD11b", "Ly6-G", "CD31", "F4_80", "Ki-67", "CD44", "CD74")

state_markers <- c("HistoneH3", "DNA1")


# Add to spe
rowData(spe)$marker_class <- ifelse(rownames(spe) %in% type_markers, "type",
                                    ifelse(rownames(spe) %in% state_markers, "state", 
                                    "other"))

## by cell type
celltype_mean <- aggregateAcrossCells(as(spe, "SingleCellExperiment"),  
                     ids = spe$celltype, 
                     statistics = "mean",
                     use.assay.type = "exprs", 
                     subset.row = rownames(spe)[rowData(spe)$marker_class == "type"]
                     )

# No scaling
dittoHeatmap(celltype_mean,
             assay = "exprs", cluster_cols = TRUE, 
             scale = "none",
             heatmap.colors = viridis(100),
             annot.by = c("celltype","ncells"))
             annotation_colors = list(celltype = metadata(spe)$color_vectors$celltype,
                                      ncells = plasma(100))
```

```{r heatmap_cellsaggreg_minmax, fig.width= 10, fig.height= 10}
# Min-max expression scaling
dittoHeatmap(celltype_mean,
             assay = "exprs", cluster_cols = TRUE, 
             scaled.to.max = TRUE,
             heatmap.colors.max.scaled = inferno(100),
             annot.by = c("celltype","ncells"))
             annotation_colors = list(celltype = metadata(spe)$color_vectors$celltype,
                                     ncells = plasma(100))
```

### Violin plot visualization

```{r violinplot_expression, fig.width= 8, fig.height= 12}
#Violin Plot - plotExpression
plotExpression(spe[,cur_cells], 
               features = rownames(spe)[rowData(spe)$marker_class == "type"],
               x = "celltype", exprs_values = "exprs", 
               colour_by = "celltype") +
    theme(axis.text.x =  element_text(angle = 90))
    scale_color_manual(values = metadata(spe)$color_vectors$celltype)
```

### Scatter plot visualization

```{r scatterplot}
#Scatter plot
dittoScatterPlot(spe, 
                 x.var = "CD3e", y.var="CD19", 
                 assay.x = "exprs", assay.y = "exprs", 
                 color.var = "celltype") +
    #scale_color_manual(values = metadata(spe)$color_vectors$celltype) +
    ggtitle("Scatterplot for CD3/CD20 labelled by celltype")
```

### Barplot visualization

```{r barplot_sampleID}
# by sample_id
dittoBarPlot(spe, var = "celltype", group.by = "ROI") 
    scale_fill_manual(values = metadata(spe)$color_vectors$celltype)
```

```{r barplot_patientID_percentage}
# by patient_id - percentage
dittoBarPlot(spe, var = "celltype", group.by = "patient_id") 
    scale_fill_manual(values = metadata(spe)$color_vectors$celltype)
```

```{r barplot_patientID_count}
# by patient_id - count
dittoBarPlot(spe, scale = "count", var = "celltype", group.by = "patient_id") 
    scale_fill_manual(values = metadata(spe)$color_vectors$celltype)
```

### CATALYST Pseudobulk-level MDS plots

```{r CATALST_setup}
# save spe in CATALYST-compatible object with renamed colData entries and 
# new metadata information
spe_cat <- spe 

spe_cat$sample_id <- factor(spe$sample_id)
spe_cat$condition <- factor(spe$condition)
spe_cat$cluster_id <- factor(spe$celltype)
spe_cat$annotation <- factor(spe$annotation)

#add celltype information to metadata
metadata(spe_cat)$cluster_codes <- data.frame(celltype = factor(spe_cat$celltype))
```

```{r MDS_plot_celltype, fig.width= 10, fig.height= 12}
# MDS pseudobulk by cell type
pbMDS(spe_cat, by = "cluster_id", 
      features = rownames(spe_cat)[rowData(spe_cat)$marker_class == "type"], 
      label_by = "cluster_id", k = "celltype") 
  scale_color_manual(values = metadata(spe_cat)$color_vectors$celltype)
```

```{r MDS_plot_celltype_condition, fig.width= 10, fig.height= 12}
# MDS pseudobulk by cell type and sample_id
pbMDS(spe_cat, by = "both", 
      features = rownames(spe_cat)[rowData(spe_cat)$marker_class == "type"], 
      k = "celltype", shape_by = "condition", 
      size_by = TRUE) 
  scale_color_manual(values = metadata(spe_cat)$color_vectors$celltype)
```

```{r MDS_plot_ann, fig.width= 12, fig.height= 10}

# MDS pseudobulk by ann

pbMDS(spe_cat, by = "sample_id", 
      color_by = "annotation", 
      features = rownames(spe_cat)[rowData(spe_cat)$marker_class == "type"]) 
  scale_color_manual(values = metadata(spe_cat)$color_vectors$ROI)
  
```

### Pseudobulk expression boxplot

```{r boxplots, fig.width= 12, fig.height= 10}
plotPbExprs(spe_cat, k = "celltype", 
            facet_by = "cluster_id", ncol = 4, 
            features = rownames(spe_cat)[rowData(spe_cat)$marker_class == "type"]) 
    scale_color_manual(values = metadata(spe_cat)$color_vectors$indication)
```

### Save data

```{r save spe3}
saveRDS(spe, file.path("output", "spe3.rds"))
```
