---
title: "IMC data analysis workflow"
author: "Tess and Tural"
date: "University of Bern, DBMR, Visceral and Transplantation Surgery lab"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Image visualization

```{r setup, include=FALSE}
#Setup knitr, incuding path
knitr::opts_chunk$set(echo = TRUE)

#Important - this path is relative to your dataset, set working directory
knitr::opts_knit$set(root.dir = "C:/Users/IMC Platform laptop/Users2/Tess/SteinbockCellSegmentation/221130_haibin/230515_workflowr_haibin")

#Set cache
knitr::opts_chunk$set(autodep = TRUE, cache = TRUE, cache.lazy = FALSE)

#To make figures
knitr::opts_chunk$set(dev = "png")
```

```{css, echo=FALSE}
.scroll-100 {
  max-height: 100px;
  overflow-y: auto;
  background-color: inherit;
}
```

### Load packages 

```{r load-libs, message=FALSE}
library(SpatialExperiment)
library(cytomapper)
library(cowplot)
library(gridGraphics)
```

### load data

```{r load data}

spe <- readRDS(file.path("output", "spe3.rds"))
images <- readRDS("data/images.rds")
masks <- readRDS("data/masks.rds")


# Sample cells
#cur_cells <- sample(seq_len(ncol(spe)), 2000)

# Sample images 

cur_id <- sample(unique(spe$sample_id), 5)
cur_images <- images[names(images) %in% cur_id]
cur_masks <- masks[names(masks) %in% cur_id]

```

### Pixel visualization: Single target expression

```{r tissue_target_ecad, fig.width=12, fig.height=12}
plotPixels(cur_images, 
           colour_by = "E-cadherin",
           bcg = list("E-cadherin" = c(0, 10, 1)))
```

```{r tissue_target_panactin, fig.width=12, fig.height=12}
plotPixels(cur_images, 
           colour_by = "Pan-actin",
           bcg = list("Pan-actin" = c(0, 10, 1)))
```

```{r tissue_target_CD3, fig.width=12, fig.height=12}
plotPixels(cur_images, 
           colour_by = "CD3e",
           bcg = list("CD3e" = c(0, 20, 1)))
```

```{r tissue_target_CD19, fig.width=12, fig.height=12}
plotPixels(cur_images, 
           colour_by = "CD19",
           bcg = list("CD19" = c(0, 50, 1)))
```

```{r tissue_target_CD8, fig.width=12, fig.height=12}
plotPixels(cur_images, 
           colour_by = "CD8a",
           bcg = list("CD8a" = c(0, 10, 1)))
```

```{r tissue_target_CD4, fig.width=12, fig.height=12}
plotPixels(cur_images, 
           colour_by = "CD4",
           bcg = list("CD4" = c(0, 20, 1)))
```

```{r tissue_target_F4_80, fig.width=12, fig.height=12}
plotPixels(cur_images, 
           colour_by = "F4_80",
           bcg = list("F4_80" = c(0, 30, 1)))
```

```{r tissue_target_CD206, fig.width=12, fig.height=12}
plotPixels(cur_images, 
           colour_by = "CD206",
           bcg = list("CD206" = c(0, 20, 1)))
```

```{r tissue_target_Ki67, fig.width=12, fig.height=12}
plotPixels(cur_images, 
           colour_by = "Ki-67",
           bcg = list("Ki-67" = c(0, 10, 1)))
```

```{r tissue_target_CD44, fig.width=12, fig.height=12}
plotPixels(cur_images, 
           colour_by = "CD44",
           bcg = list("CD44" = c(0, 10, 1)))
```

```{r tissue_target_CD74, fig.width=12, fig.height=12}
plotPixels(cur_images, 
           colour_by = "CD74",
           bcg = list("CD74" = c(0, 20, 1)))
```

```{r tissue_target_aSMA, fig.width=12, fig.height=12}
plotPixels(cur_images, 
           colour_by = "a-SMA",
           bcg = list("a-SMA" = c(0, 200, 1)))
```

```{r tissue_target_CD31, fig.width=12, fig.height=12}
plotPixels(cur_images, 
           colour_by = "CD31",
           bcg = list("CD31" = c(0, 10, 1)))
```

```{r tissue_target_CD11b, fig.width=12, fig.height=12}
plotPixels(cur_images, 
           colour_by = "CD11b",
           bcg = list("CD11b" = c(0, 20, 1)))
```

```{r tissue_target_Ly6G, fig.width=12, fig.height=12}
plotPixels(cur_images, 
           colour_by = "Ly6-G",
           bcg = list("Ly6-G" = c(0, 10, 1)))
```

### Pixel visualization: Multi target expression

```{r tissue_targets_multi1, fig.width=12, fig.height=12}
plotPixels(cur_images, 
           colour_by = c("E-cadherin","CD206", "CD3e", "CD19", "CD31"),
           bcg = list("E-cadherin" = c(0, 10, 1),
                      CD206 = c(0, 20, 1),
                      CD3e = c(0, 20, 1),
                      CD19 = c(0, 50, 1),
                      CD31 = c(0, 10, 1)))
```

```{r Panactin_F480, fig.width=12, fig.height=12}
plotPixels(cur_images, 
           colour_by = c("Pan-actin","F4_80", "DNA2"),
           bcg = list("Pan-actin" = c(0, 8, 1),
                      "F4_80" = c(0, 30, 1),
                      DNA2 = c(0, 10, 1)))
```

```{r Panactin_CD206, fig.width=12, fig.height=12}
plotPixels(cur_images, 
           colour_by = c("Pan-actin","CD206", "DNA2"),
           bcg = list("Pan-actin" = c(0, 8, 1),
                      "CD206" = c(0, 20, 1),
                      DNA2 = c(0, 10, 1)))
```

```{r Panactin_CD44, fig.width=12, fig.height=12}
plotPixels(cur_images, 
           colour_by = c("Pan-actin","CD44", "DNA2"),
           bcg = list("Pan-actin" = c(0, 8, 1),
                      "CD44" = c(0, 10, 1),
                      DNA2 = c(0, 10, 1)))
```

```{r Panactin_Ki67, fig.width=12, fig.height=12}
plotPixels(cur_images, 
           colour_by = c("Pan-actin","Ki-67", "DNA2"),
           bcg = list("Pan-actin" = c(0, 8, 1),
                      "Ki-67" = c(0, 10, 1),
                      DNA2 = c(0, 10, 1)))
```

```{r Panactin_CD11b, fig.width=12, fig.height=12}
plotPixels(cur_images, 
           colour_by = c("Pan-actin","CD11b", "DNA2"),
           bcg = list("Pan-actin" = c(0, 8, 1),
                      "CD11b" = c(0, 20, 1),
                      DNA2 = c(0, 10, 1)))
```

```{r Panactin_ecad, fig.width=12, fig.height=12}
plotPixels(cur_images, 
           colour_by = c("Pan-actin","E-cadherin", "DNA2"),
           bcg = list("Pan-actin" = c(0, 8, 1),
                      "E-cadherin" = c(0, 10, 1),
                      DNA2 = c(0, 10, 1)))
```

```{r tissue_targets_colorchoose, fig.width=12, fig.height=12}
plotPixels(cur_images, 
           colour_by = c("CD206", "Pan-actin", "E-cadherin"),
           bcg = list(CD206 = c(0, 20, 1),
                      "Pan-actin" = c(0, 8, 1),
                      "E-cadherin" = c(0, 10, 1)),
           colour = list(CD206 = c("black", "burlywood1"),
                         "Pan-actin" = c("black", "cyan2"),
                         "E-cadherin" = c("black", "firebrick1")))
```

### Pixel visualization: different normalizations of images

```{r tissue_targets_normed_acrossall, eval=FALSE, include=FALSE}
# 0 - 1 channel scaling across all images
norm_images <- normalize(cur_images)

# Clip channel at 0.2
#norm_images <- normalize(norm_images, inputRange = c(0, 0.2))

plotPixels(norm_images, 
           colour_by = c("E-cadherin","CD206", "HNFa", "C16", "CD19", "CD31"))
```

```{r tissue_targets_normed_perimage, eval=FALSE, include=FALSE}
# 0 - 1 channel scaling per image
norm_images <- normalize(cur_images, separateImages = TRUE)

# Clip channel at 0.2
norm_images <- normalize(norm_images, inputRange = c(0, 0.2))

plotPixels(norm_images, 
           colour_by = c("E-cadherin","CD206", "HNFa", "16", "CD19", "CD31"))
```

```{r tissue_targets_normed_manual, eval=FALSE, include=FALSE}
# 0 - 1 channel scaling per image
norm_images <- normalize(cur_images, 
                         separateImages = TRUE,
                         inputRange = list("E-cadherin" = c(0, 10), 
                                           CD206 = c(0, 10),
                                           HNFa = c(0, 10),
                                           "16" = c(0, 10),
                                           CD19 = c(0, 10),
                                           CD31 = c(0, 10)))

plotPixels(norm_images, 
           colour_by = c("E-cadherin","CD206", "HNFa", "16", "CD19", "CD31"))
```

### Cell type overlays

```{r celltype_overlay}
plotCells(cur_masks,
          object = spe, 
          cell_id = "ObjectNumber", img_id = "sample_id",
          colour_by = "celltype")
```

```{r celltype_overlay_choosecolors}
plotCells(cur_masks,
          object = spe, 
          cell_id = "ObjectNumber", img_id = "sample_id",
          colour_by = "celltype")
          colour = list(celltype = metadata(spe)$color_vectors$celltype)
```

```{r celltype_overlay_onetarget_tumor_red}

tumor <- spe[,spe$celltype == "Tumor"]

plotCells(cur_masks,
          object = tumor, 
          cell_id = "ObjectNumber", img_id = "sample_id",
          colour_by = "celltype",
          colour = list(celltype = c(Tumor = "red")),
          missing_colour = "white")
```

### Cell area overlay

```{r area_overlay}
plotCells(cur_masks,
          object = spe, 
          cell_id = "ObjectNumber", img_id = "sample_id",
          colour_by = "area")
```

### Asinh-transformed mean pixel intensities

```{r meanpixel_overlay_onetarget, fig.width=12, fig.height=12}
plotCells(cur_masks,
          object = spe, 
          cell_id = "ObjectNumber", img_id = "sample_id",
          colour_by = "E-cadherin",
          exprs_values = "exprs")
```

```{r meanpixel_overlay_multitarget, fig.width=12, fig.height=12}
plotCells(cur_masks,
          object = spe, 
          cell_id = "ObjectNumber", img_id = "sample_id",
          colour_by = c("E-cadherin","CD206", "Pan-actin", "CD3e", "CD19", "Ki-67"),
          exprs_values = "exprs")
```

```{r meanpixel_overlay_multitarget_manualcolors, fig.width=12, fig.height=12}
plotCells(cur_masks,
          object = spe, 
          cell_id = "ObjectNumber", img_id = "sample_id",
          colour_by = c("E-cadherin","CD206", "Pan-actin"),
          exprs_values = "exprs",
          colour = list("E-cadherin" = c("black", "burlywood1"),
                          CD206 = c("black", "cyan2"),
                          "Pan-actin" = c("black", "firebrick1")))
```

### Outlining cells on images

```{r celloutline_overlay_all}
plotPixels(image = cur_images,
           mask = cur_masks,
           object = spe, 
           cell_id = "ObjectNumber", img_id = "sample_id",
           colour_by = c("E-cadherin","CD206", "DNA1"),
           outline_by = "celltype",
                      bcg = list("E-cadherin" = c(0, 10, 1),
                      CD206 = c(0, 20, 1),
                      DNA1 = c(0, 10, 1)),
           #colour = list(celltype = metadata(spe)$color_vectors$celltype),
           thick = TRUE)
```

```{r celloutline_overlay_onecelltype}

CD8 <- spe[,spe$celltype == "CD8"]

plotPixels(image = cur_images,
           mask = cur_masks,
           object = CD8, 
           cell_id = "ObjectNumber", img_id = "sample_id",
           colour_by = c("CD3e", "CD8a"),
           outline_by = "celltype",
                      bcg = list(CD3e = c(0, 20, 1),
                      CD8a = c(0, 20, 1)),
           colour = list(celltype = c("CD8" = "white")),
           thick = TRUE)
```

### Adjusting plot annotation

```{r target_overlay_annotated, fig.width=13, fig.height=10, fig.width=12, fig.height=12}

plotPixels(cur_images, 
           colour_by = c("E-cadherin","CD206", "CD3e", "CD19", "CD31"),
           bcg = list("E-cadherin" = c(0, 10, 1),
                      CD206 = c(0, 20, 1),
                      CD3e = c(0, 20, 1),
                      CD19 = c(0, 50, 1),
                      CD31 = c(0, 10, 1)),
           scale_bar = list(length = 100,
                            label = expression("100 " ~ mu * "m"),
                            cex = 0.7, 
                            lwidth = 10,
                            colour = "grey",
                            position = "bottomleft",
                            margin = c(5,5),
                            frame = 3),
           image_title = list(text = mcols(cur_images)$condition,
                              position = "topright",
                              colour = "grey",
                              margin = c(5,5),
                              font = .5,
                              cex = 2),
           legend = list(colour_by.title.cex = 0.7,
                         margin = 10),
           margin = 40)
```

### Display individual images

```{r celltype_overlay_oneimage}
plotCells(cur_masks,
          object = spe, 
          cell_id = "ObjectNumber", img_id = "sample_id",
          colour_by = "celltype",
          #colour = list(celltype = metadata(spe)$color_vectors$celltype),
          display = "single",
          legend = NULL)
```

### Save and return images

```{r, save and return images, eval=FALSE, include=FALSE}
out1 <- plotCells(cur_masks,
                  object = spe, 
                  cell_id = "ObjectNumber", img_id = "sample_id",
                  colour_by = "celltype",
                  #colour = list(celltype = metadata(spe)$color_vectors$celltype),
                  return_plot = TRUE)

out2 <- plotCells(cur_masks,
                  object = spe, 
                  cell_id = "ObjectNumber", img_id = "sample_id",
                  colour_by = c("Pan-actin", "CD3e", "CD206"),
                  exprs_values = "exprs",
                  return_plot = TRUE)

p1 <- ggdraw(out1$plot, clip = "on")
p2 <- ggdraw(out2$plot, clip = "on")

plot_grid(p1, p2)
```
